Open Policing Project Tutorial

### 2019-03-06 nicar_tutorial.Rmd
incorporate hans' edits

---
title: "Open Policing Project Tutorial"
output: html_document
---

@@

## Load the data
# Replace the path below with the path to where your data lives
# data_path <- "~/opp/data/states/pa/philadelphia/clean/philadelphia.rds"
# stops <- read_rds(data_path)$data
data_path <- "~/Desktop/pa_philadelphia_2019_02_25.rds"
stops <- read_rds(data_path)

@@

@@

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE
colnames(stops)
```

How many stops do we have in our dataset?

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE
nrow(stops)
```

What date range does our data cover?

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE
min(stops$date)
max(stops$date)
```

@@

You might have noticed in the colnames, that we have no `year` column (our data 
only has `date`). Luckily, with the `lubridate` package, getting the year from
the date is as easy as `year(date)`!

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE
stops <- stops %>% filter(year(date) < 2018)
```

@@

**tidyverse tip:** The "pipe" operator, `%>%`, just places the previous item
into the first argument of the function. So, `x %>% f()` is simply `f(x)`. While
in a one-function call the pipe might feel silly and unnecessary, it's going to
become _really_ helpful once we start wanting to do multiple transformations to
our data.

@@

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE
stops %>% 
  select(type) %>% 
  head(10) # this allows us to look at just the first 10 rows
```

@@

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE
stops <- stops %>% filter(type == "vehicular")
```

@@

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE
stops %>% nrow()
```

@@

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE
stops %>%
  mutate(year = year(date)) %>% 
  count(year)
```

@@

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE
stops %>% 
  count(subject_race)
```

@@

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE

# This method builds off of using `count` as above
stops %>% 
  count(subject_race) %>% 
  mutate(prop = n / sum(n))

@@

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE

stops %>% 
  # notice that you can also mutate `date` *within* the count funciton
  count(year = year(date), subject_race)
```

@@

There are many tweaks we could make to this plot to make it nicer, but in the 
phase of exploring the data, it's enough just to get a coarse visual like this.

From this plot we see that, at least for black and white drivers (it's hard to
tell from this plot for drivers of other races because the counts are 
comparatively so much smaller), the annual trends are very different by race.
All races experienced a spike in enforcement in 2015, but thereafter, there
were fewer white drivers stopped (in 2016 and 2017), whereas there continued to
be an _increase_ in number of black drivers stopped over those two years.

This is already a potential lead! We'd have to investigate further to see what 
the trend looks like when adjusting for population changes over time, etc. But 
it's often simple explorations like this that can uncover the path to stories.

@@

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE

stops <- stops %>% filter(year(date) == 2017)
```

@@

In order to do this baseline comparison, we need to understand the racial
demographics in our Philly population data. Using the `population_2017` data 
we've given you, compute the proportions by race.

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE

population_2017 %>% 
  mutate(prop = num_people / sum(num_people))
```

@@

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE

stops %>% 
  count(subject_race) %>% 
  left_join(
    population_2017,
    by = "subject_race"
  ) %>% 
  mutate(stop_rate = n / num_people)
```

Good! Now this has allows us to make the quantitative claim that blacks are
stopped at a rate 3.4 times higher than whites are, and Hispanics are stopped at
a rate 1.5 times higher than whites are.

@@

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE

stops %>% 
  group_by(subject_race) %>% 
  summarize(
    search_rate = mean(search_conducted, na.rm = T),
    frisk_rate = mean(frisk_performed, na.rm = T)
  )
```

Here we see that among drivers who were stopped, blacks were searched at a rate
1.5 times higher than whites were, and Hispanics were searched at a rate 1.2
times higher than whites were. Black drivers were frisked at a rate 2.1 times
higher than white drivers were, and Hispanic drivers were frisked at a rate
1.5 times higher than white drivers were.

@@

As a caricatured example, suppose among white drivers who were searched, 
officers found contraband 99% of the time, while among black drivers who were
searched, officers found contraband only 1% of the time. This would lead us to
believe that officers made sure they were _certain_ white individuals had
contraband before deciding to search, but that they were searching black 
individuals on a whif of evidence.

@@

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE

stops %>% 
  filter(search_conducted) %>% 
  group_by(subject_race) %>% 
  summarize(
    hit_rate = mean(contraband_found, na.rm = T)
  )
```

@@

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE

hit_rates <- 
  stops %>% 
  filter(search_conducted) %>% 
  group_by(subject_race, district) %>% 
  summarize(hit_rate = mean(contraband_found, na.rm = T))
```

Again, this is too many hit rates to compare in one table. To plot the hit rates
of blacks vs whites and of Hispanics vs whites, we need to reshape our table
to have each row containing a district, a minority race, minority hit rate in
that district, and white hit rate in that district. We'll walk you through the 
code below that reshapes the data for us.

@@

@@

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE

hit_rates %>% 
  filter(white_hit_rate == min(white_hit_rate))
```

@@

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE

@@

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE
# Hint: Note that districts in our dataset are encoded as characters, 
# so 77 is actually "77"

@@

```{r}
### Remove for tutorial, replace with:
# YOUR CODE HERE
sunset_times %>% 
  filter(dusk == min(dusk) | dusk == max(dusk))
```

@@

* filter to the inter-twiligh period
* remove that ambigous pre-dusk period
* for simplicity, compare only black and white drivers

@@

~~
**NOTE FOR ERIC/JUSTIN/CHERYL/BETHANY/HANS:** I can add in another raw data intuition
section that's a basic raw data plot if you think the concept of the VOD test
is still confusing. But i want to keep things a bit briefer that way we have 
more time to field questions and have them actually type things in and think 
things through rather than just running code dan and i wrote...let me know what
you think though.
~~

>>>>

---
title: "Open Policing Project Tutorial"
output:
  html_document:
    code_folding: hide
---

@@

## Load the data
# Replace the path below with the path to where your data lives
data_path <- "~/opp/data/states/pa/philadelphia/clean/philadelphia.rds"
stops <- read_rds(data_path)$data
# data_path <- "~/Desktop/pa_philadelphia_2019_02_25.rds"
# stops <- read_rds(data_path)

@@

The core of `R` is the dataframe. We've given you one to start with, in the form
of `stops`. Think of dataframes like a spreadsheet: they have rows and columns.
Usually, rows are a "datapoint": in `stops`, each row corresponds to a single
stop from Philadelphia. The columns are the "variables": again, in `stops`,
these are the things we know about the stop, like where the stop happened, the
age of the stop subject, whether an arrest was made, and so on.

@@

```{r}
colnames(stops)
```

How many stops do we have in our dataset? (**Hint**: Try the `nrow()` function.)

```{r}
nrow(stops)
```

What date range does our data cover? One way to do this would be to examine the
`min()` and `max()` dates in our dataframe. The problem is we can't just apply
`min()` to our dataframe: doing so produces an error.

The `min()` function wants its input to be just one column from the `stops`
dataframe. To access a single column from a dataframe, use `$`: what goes before
`$` is the dataframe, and what comes after is the name of the column we want.
(**Hint**: What happens if you type `stops$date` into the console?)

```{r}
min(stops$date)
max(stops$date)
```

@@

#### Aside: `filter()`

The `filter()` function is used to separate rows from the dataframe that
interest us from rows that do not. In particular, `filter(DATA, CONDITION)`
returns `DATA` with all of the rows that don't satisfy `CONDITION` removed. For
instance, we might want to only look at stops where the subject was arrested. To
do this, we would type `filter(stops, arrest_made == TRUE)`, since we only want
rows from `stops` where the `arrest_made` column is (i.e., `==`) `TRUE`. (Note:
we could actually just type `filter(stops, arrest_made)`. Do you see why?)

You might have noticed in the colnames, that we have no `year` column (our data 
only has `date`). These dates are full dates that look like `YYYY-MM-DD`.
Luckily, with the `lubridate` package, getting the year from
the date is as easy as `year(date)`!

```{r}
stops <- stops %>% filter(year(date) < 2018)
```

@@

**tidyverse tip:** The "pipe" operator, `%>%`, helps to keep our code clean. It
just places the previous item into the first argument of the function. So,
`x %>% f()` is simply `f(x)`. While in a one-function call the pipe might feel
silly and unnecessary, it's going to become _really_ helpful once we start
wanting to do multiple transformations to our data. For instance, that means
that instead of typing `nrows(filter(stops, year < 2018))` we can write
`stops %>% filter(year < 2018) %>% nrows()`, which is easier to understand.

@@

#### Aside: `select()`
A few minutes ago, we used `filter()` to whittle our dataframe down to just the
rows that interested us. What happens if we want to whittle down the columns
instead? The `select()` function is designed to solve exactly this problem. It
takes a dataframe, and gets rid of all of the columns you don't ask for. For
instance, `select(stops, lat, long)` produces a dataframe with exactly two
columns: `lat` and `long`. If we wanted to know the time as well, we could type
`select(stops, lat, long, time)`.

Try using `select()` to look at the `type` column.

```{r}
stops %>% 
  select(type) %>% 
  head(10) # this allows us to look at just the first 10 rows
```

@@

```{r}
stops <- stops %>% filter(type == "vehicular")
```

@@

```{r}
stops %>% nrow()
```

@@

#### Aside: `mutate()`

Notice, though, that we have the same problem as before: there's no `year`
column in the `stops` dataframe. We'll use the `mutate()` function to fix that.
The `mutate()` function adds new columns to a dataframe based on old columns.
The basic setup is `mutate(DATA, NEW_COL = FUN(OLD_COL))` where `DATA` is our
dataframe, `NEW_COL` is the name of the new column we want, and `FUN` is the
function we apply to the old column, `OLD_COL`, to get it.

Use the `year()` and `mutate()` functions to add a new column called `year`
to our `stops` dataframe and then use `count()` to determine the number of stops per year.

```{r}
stops %>%
  mutate(year = year(date)) %>% 
  count(year)
```

@@

```{r}
stops %>% 
  count(subject_race)
```

@@

```{r}
# This method builds off of using `count` as above
stops %>% 
  count(subject_race) %>% 
  mutate(prop = n / sum(n))

@@

```{r}
stops %>% 
  # notice that you can also mutate `date` *within* the count funciton
  count(year = year(date), subject_race)
```

@@

We used ``ggplot`` to make this plot. This package provides a lot of options for
customization, and there are many tweaks we could make to this plot to make it
nicer. It's syntax is a little complicated, though, and in our data
exploration, it's enough just to get a coarse visual like this.

From this plot we see that, at least for black and white drivers, the annual
trends are very different by race. (It's hard to tell from this plot for
drivers of other races because the counts are comparatively so much smaller.) 
All races experienced a spike in enforcement in 2015, but thereafter, there
were fewer white drivers stopped (in 2016 and 2017), whereas there continued to
be an _increase_ in number of black drivers stopped over those two years.

This is already a potential lead! We'd have to investigate further to see what 
the trend looks like when adjusting for population changes over time or other
factors we haven't considered in this example. But it's often simple
explorations like this that can uncover the path to stories.

@@

```{r}
stops <- stops %>% filter(year(date) == 2017)
```

@@

In order to do this baseline comparison, we need to understand the racial
demographics in our Philly population data. The data as we've given it to you
has raw population numbers. To make it useful, we'll need to compute the
_proportion_ of Philadelphia residents in each demographic group. (Hint: use the
`mutate()` function.)

```{r}
population_2017 %>% 
  mutate(prop = num_people / sum(num_people))
```

@@

#### Aside: `left_join()`

The way we put tables together is with the `left_join()` function. We need to
input three things into this function: the two tables we'd like to combine,
along with instructions on how to join them. In this case, the two tables we
want to join is the table of stops counted according to `subject race` and
`population_2017`. The instruction for combining the tables is to join rows that
contain the same race, so we have to also give `left_join()` the argument `by =
"subject_race"`. This means that `left_join()` will look at the first table,
i.e., the table stops counted by race, and then go to the `subject_race` column
in each row. Then, `left_join()` will take what it finds there---in this case,
`"asian/pacific islander"`, `"black"`, `"hispanic"`, `"other/unknown"`, and
`"white"`---and look in the second table, i.e., `population_2017`, for all the
rows that contain the same information in `population_2017`'s race column. Then,
it will add the second row at the end of the first to create a new row with
information from both. What we end up with is a dataframe with all of the
columns from __both__ tables.

The process is a little complicated, and we won't use it too much, so don't
worry if the abstract description doesn't make sense. To get a better
understanding of what's going on, try joining the two tables described above,
being sure to include the `by = "subject_race"` argument.

```{r}
stops %>% 
  count(subject_race) %>% 
  left_join(
    population_2017,
    by = "subject_race"
  ) %>% 
  mutate(stop_rate = n / num_people)
```

Good! Now we can divide the black stop rate by the white stop rate to be able to make 
a quantitative statement about how much more often blacks are stops versus whites,
relative to their share of the city's population. Blacks are stopped at a rate 3.4 
times higher than whites are, and Hispanics are stopped at a rate 1.5 times higher 
than whites are.

@@

#### Aside: `group_by()` and `summarize()`

One thing that we often want to do with data is disaggregate it. That is, we
might want to take the data and break it down into smaller subpopulations. Then,
when we ask questions, we can ask about each piece---for instance, each
demographic group, or each police district---instead asking about the population
as a whole.

The way to do this in `R` is with `group_by()` and summarise. The standard way
to use `group_by()` is to call `group_by(DATA, COL_NAME)`, where `DATA` is our
dataframe and `COL_NAME` is the name of a column. What `group_by()` then does is
take all the rows in the dataframe `DATA` and put them into different groups,
one for each different value in the column `COL_NAME`. So, for instance, if we
called `group_by(stops, district)`, `R` would hand back to us the `stops`
dataframe with all of its columns broken into different groups, one for each
police district.

The second step is to do something with those groups. That's what `summarize()`
does. The way `summarize()` works is, given a dataframe broken into groups by
`group_by()`, will calculate a value, one for each group. The basic syntax is
`summarize(DATA, STAT = FUN(COL_NAME))`, where `DATA` is some dataframe broken
up by `group_by()`, `STAT` is some statistic we want to calculate, `FUN` is the
function that calculates that statistic, and `COL_NAME` is the name of the
column (or columns) used to calculate the statistic.

Let's put it all together with an example. A straightforward one is: `stops %>%
group_by(subject_race) %>% summarize(arrest_rate = mean(arrest_made, na.rm =
True))`. Typing this into the `R` console will calculates arrest rates
disaggregated by race.

Using `group_by()` and `summarize()`, let's calculate frisk rates by race.

```{r}
stops %>% 
  group_by(subject_race) %>% 
  summarize(
    search_rate = mean(search_conducted, na.rm = T),
    frisk_rate = mean(frisk_performed, na.rm = T)
  )
```

As with the stop rates, we can make a quantitative claim about disparities in
search and frisk rates by dividing the minority rate by the white rate.
Here we see that among drivers who were stopped, blacks were searched at a rate
1.5 times higher than whites were, and Hispanics were searched at a rate 1.2
times higher than whites were. Black drivers were frisked at a rate 2.1 times
higher than white drivers were, and Hispanic drivers were frisked at a rate
1.5 times higher than white drivers were.

@@

As a caricatured example, suppose among white drivers who were searched, 
officers found contraband 99% of the time, while among black drivers who were
searched, officers found contraband only 1% of the time. This would lead us to
believe that officers made sure they were _certain_ white individuals had
contraband before deciding to search, but that they were searching black 
individuals on a whiff of evidence.

@@

```{r}
stops %>% 
  filter(search_conducted) %>% 
  group_by(subject_race) %>% 
  summarize(
    hit_rate = mean(contraband_found, na.rm = T)
  )
```

@@

```{r}
hit_rates <- 
  stops %>% 
  filter(search_conducted) %>% 
  group_by(subject_race, district) %>% 
  summarize(hit_rate = mean(contraband_found, na.rm = T))
```

Again, this is too many hit rates to compare in one table. To plot the hit rates
of blacks vs. whites and of Hispanics vs. whites, we need to reshape our table
to have each row containing a district, a minority race, minority hit rate in
that district, and white hit rate in that district. We'll walk you through the 
code below that reshapes the data for us.

@@

**Quick Tip**: You might have noticed a strange symbol in the code above:
`%in%`. This is really just shorthand for the sentence, "the `subject_race` is
`"black"` or its `"white"` or its `"hispanic"`." Without the `%in%` syntax, we
would have to write this in `R` as: `subject_race == "black" | subject_race ==
"white" | subject_race == "hispanic"`. The `%in%` syntax is a lot shorter and
clearer, we think you'll agree.

@@

```{r}
hit_rates %>% 
  filter(white_hit_rate == min(white_hit_rate))
```

@@

```{r}
# Hint: Note that districts in our dataset are encoded as characters, 
# so 77 is actually "77"

@@

```{r}
# Remember: Districts in our dataset are encoded as characters, 
# so 77 is actually "77"

@@

```{r}
sunset_times %>% 
  filter(dusk == min(dusk) | dusk == max(dusk))
```

@@

* filter to the inter-twilight period
* remove that ambigous pre-dusk period
* for simplicity, compare only black and white drivers

@@
